---
title: "Exploratory Data Analysis (EDA) - IEEE-CIS Fraud Detection"
author: "Manuel Alejandro Matías Astorga"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: false
    theme: flatly
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 10,
  fig.height = 6
)
```

## 1. Introduction

This notebook presents an Exploratory Data Analysis (EDA) of the [IEEE-CIS Fraud Detection](https://www.kaggle.com/competitions/ieee-fraud-detection) dataset, provided by Vesta Corporation for a Kaggle competition. The dataset contains anonymized transactional and identity information with the the goal of detecting fraudulent transactions.

**Key questions this EDA answers:**

- How severe is the class imbalance, and what does it imply for modeling?
- What patterns in transaction amount, time, and device distinguish fraud from legitimate transactions?
- Which features carry the most signal for fraud detection?
- What multicollinearity exists among predictors?
- Where are the missing values concentrated, and are they informative?

## 2. Libraries

We'll require the next libraries for data manipulation and visualization.

```{r libraries}
library(tidyverse)
library(arrow) # read parquet — replaces data.table for I/O
library(gridExtra)
library(skimr)
library(ggcorrplot)
library(naniar)
```

## 3. Data Loading

The dataset is composed of two main files:

- **`train_transaction`**: transactional data for online purchases — amount,
  product codes, card and address information, and the fraud label (`isFraud`).

- **`train_identity`**: identity-related features (device info, browser, email
  domain) for a subset of transactions. Only ~24% of transactions have
  corresponding identity information.

To perform a comprehensive analysis, both datasets are merged using
`TransactionID` as the join key. A **left join** preserves all transactions
from `train_transaction`, including those without identity information.
This ensures:

- The **full set of transactions** is available for fraud analysis.
- Identity features are included **only when available** — their absence
  is itself informative and will be considered during feature engineering.

```{r data-loading}
train_transaction <- read_parquet(
  "../../data/processed/train_transaction.parquet"
)
train_identity <- read_parquet(
  "../../data/processed/train_identity.parquet"
)

cat(
  "train_transaction:", nrow(train_transaction), "rows,",
  ncol(train_transaction), "cols\n"
)
cat(
  "train_identity:   ", nrow(train_identity), "rows,",
  ncol(train_identity), "cols\n"
)
```

```{r data-join}
train <- train_transaction |>
  left_join(train_identity, by = "TransactionID")

cat("Combined dataset:", nrow(train), "rows,", ncol(train), "cols\n")
```

`train_transaction` contains 590,540 rows and 394 columns, while
`train_identity` has 144,233 rows and 41 columns. The merged dataset
has **590,540 rows and 434 columns** — `TransactionID` appears in both
files but is not duplicated after the join.

Due to the high dimensionality and anonymized nature of the dataset
(434 features), this EDA focuses on interpretable and structurally
relevant features: `TransactionAmt`, `DeviceType`, email domains,
`C1–C14`, and `D1–D15`. Feature importance and selection will be
addressed in the modeling stage using SHAP values.

## 4. Data Overview

Let's analyze the dataset to understand its structure, missing values, and class balance.

### 4.1 Missing Values

To assess the quality of the data, we calculated the percentage of missing values in each column using the `summarise_all(~ mean(is.na(.)))` function chain. This provides a quick overview of how much information is missing in each feature.

```{r missing-values}
missing_data <- train |>
  summarise(across(everything(), ~ mean(is.na(.)))) |>
  pivot_longer(
    cols = everything(),
    names_to = "variable",
    values_to = "missing_perc"
  ) |>
  arrange(desc(missing_perc))

# View top 20 variables with the highest percentage of missing values
head(missing_data, 20)
```

From the results above, we observe that several variables (especially `id_24`, `id_25`, `id_07`, `id_08`, and many features from the `id_` and `D` families) have over 85–99% missing values. This is critical because:

- Such high missingness may suggest that these features are either optional, sparsely recorded, or system-dependent.
- These features may not be useful for modeling without appropriate imputation, encoding, or dimensionality reduction.
- We must decide whether to drop these variables or retain them by filling missing values (e.g., using domain knowledge or imputation strategies).

This information will be key during the feature selection and preprocessing phases.

### 4.2 Class Balance

This dataset is highly imbalanced, with a strong predominance of non-fraudulent transactions. Understanding this imbalance is crucial before training machine learning models, as it can bias the models toward the majority class.

```{r class-balance}
# Distribution of target variable
train |>
  count(isFraud) |>
  mutate(percentage = n / sum(n))
```

**Observation**:

The target variable `isFraud` is heavily imbalanced:

- **Non-fraudulent transactions** (isFraud = 0): 569,877 (~96.5%)
- **Fraudulent transactions** (isFraud = 1): 20,663 (~3.5%)

This imbalance has direct implications for modeling and evaluation:

- **Accuracy is misleading**: a model predicting "no fraud" for every
  transaction achieves 96.5% accuracy while detecting zero fraud cases.
  AUC-ROC and Average Precision (AUC-PR) will be used instead.

- **Class weighting over resampling**: `scale_pos_weight` in LightGBM
  penalizes misclassification of the minority class without generating
  synthetic data. Resampling techniques like SMOTE are avoided —
  generating synthetic fraudulent transactions in a 400+ dimensional
  space produces samples that do not reflect real fraud patterns.

- **Threshold tuning**: the default classification threshold of 0.5
  is suboptimal for imbalanced datasets. Since this is a Kaggle
  competition evaluated on AUC-ROC — a threshold-free metric —
  threshold selection is not critical here, but would be a key
  decision in a production deployment where false positive and
  false negative costs differ.
