---
title: "Exploratory Data Analysis (EDA) - IEEE-CIS Fraud Detection"
author: "Manuel Alejandro Matías Astorga"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: false
    theme: flatly
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 10,
  fig.height = 6
)
```

## 1. Introduction

This notebook presents an Exploratory Data Analysis (EDA) of the [IEEE-CIS Fraud Detection](https://www.kaggle.com/competitions/ieee-fraud-detection) dataset, provided by Vesta Corporation for a Kaggle competition. The dataset contains anonymized transactional and identity information with the the goal of detecting fraudulent transactions.

**Key questions this EDA answers:**

- How severe is the class imbalance, and what does it imply for modeling?
- What patterns in transaction amount, time, and device distinguish fraud from legitimate transactions?
- Which features carry the most signal for fraud detection?
- What multicollinearity exists among predictors?
- Where are the missing values concentrated, and are they informative?

## 2. Libraries

We'll require the next libraries for data manipulation and visualization.

```{r libraries}
library(tidyverse)
library(arrow) # read parquet — replaces data.table for I/O
library(gridExtra)
library(skimr)
library(ggcorrplot)
library(naniar)
```

## 3. Data Loading

The dataset is composed of two main files:

- **`train_transaction`**: transactional data for online purchases — amount,
  product codes, card and address information, and the fraud label (`isFraud`).

- **`train_identity`**: identity-related features (device info, browser, email
  domain) for a subset of transactions. Only ~24% of transactions have
  corresponding identity information.

To perform a comprehensive analysis, both datasets are merged using
`TransactionID` as the join key. A **left join** preserves all transactions
from `train_transaction`, including those without identity information.
This ensures:

- The **full set of transactions** is available for fraud analysis.
- Identity features are included **only when available** — their absence
  is itself informative and will be considered during feature engineering.

```{r data-loading}
train_transaction <- read_parquet(
  "../../data/processed/train_transaction.parquet"
)
train_identity <- read_parquet(
  "../../data/processed/train_identity.parquet"
)

cat(
  "train_transaction:", nrow(train_transaction), "rows,",
  ncol(train_transaction), "cols\n"
)
cat(
  "train_identity:   ", nrow(train_identity), "rows,",
  ncol(train_identity), "cols\n"
)
```

```{r data-join}
train <- train_transaction |>
  left_join(train_identity, by = "TransactionID")

cat("Combined dataset:", nrow(train), "rows,", ncol(train), "cols\n")
```

`train_transaction` contains 590,540 rows and 394 columns, while
`train_identity` has 144,233 rows and 41 columns. The merged dataset
has **590,540 rows and 434 columns** — `TransactionID` appears in both
files but is not duplicated after the join.

Due to the high dimensionality and anonymized nature of the dataset
(434 features), this EDA focuses on interpretable and structurally
relevant features: `TransactionAmt`, `DeviceType`, email domains,
`C1–C14`, and `D1–D15`. Feature importance and selection will be
addressed in the modeling stage using SHAP values.

## 4. Data Overview

Let's analyze the dataset to understand its structure, missing values, and class balance.

### 4.1 Missing Values

To assess the quality of the data, we calculated the percentage of missing values in each column using the `summarise_all(~ mean(is.na(.)))` function chain. This provides a quick overview of how much information is missing in each feature.

```{r missing-values}
missing_data <- train |>
  summarise(across(everything(), ~ mean(is.na(.)))) |>
  pivot_longer(
    cols = everything(),
    names_to = "variable",
    values_to = "missing_perc"
  ) |>
  arrange(desc(missing_perc))

# View top 20 variables with the highest percentage of missing values
head(missing_data, 20)
```

From the results above, we observe that several variables (especially `id_24`, `id_25`, `id_07`, `id_08`, and many features from the `id_` and `D` families) have over 85–99% missing values. This is critical because:

- Such high missingness may suggest that these features are either optional, sparsely recorded, or system-dependent.
- These features may not be useful for modeling without appropriate imputation, encoding, or dimensionality reduction.
- We must decide whether to drop these variables or retain them by filling missing values (e.g., using domain knowledge or imputation strategies).

This information will be key during the feature selection and preprocessing phases.

### 4.2 Class Balance

This dataset is highly imbalanced, with a strong predominance of non-fraudulent transactions. Understanding this imbalance is crucial before training machine learning models, as it can bias the models toward the majority class.

```{r class-balance}
# Distribution of target variable
train |>
  count(isFraud) |>
  mutate(percentage = n / sum(n))
```

**Observation**:

The target variable `isFraud` is heavily imbalanced:

- **Non-fraudulent transactions** (isFraud = 0): 569,877 (~96.5%)
- **Fraudulent transactions** (isFraud = 1): 20,663 (~3.5%)

This imbalance has direct implications for modeling and evaluation:

- **Accuracy is misleading**: a model predicting "no fraud" for every
  transaction achieves 96.5% accuracy while detecting zero fraud cases.
  AUC-ROC and Average Precision (AUC-PR) will be used instead.

- **Class weighting over resampling**: `scale_pos_weight` in LightGBM
  penalizes misclassification of the minority class without generating
  synthetic data. Resampling techniques like SMOTE are avoided —
  generating synthetic fraudulent transactions in a 400+ dimensional
  space produces samples that do not reflect real fraud patterns.

- **Threshold tuning**: the default classification threshold of 0.5
  is suboptimal for imbalanced datasets. Since this is a Kaggle
  competition evaluated on AUC-ROC — a threshold-free metric —
  threshold selection is not critical here, but would be a key
  decision in a production deployment where false positive and
  false negative costs differ.

## 5. Exploratory Visualizations

### 5.1 Transaction Amount Distribution

```{r transaction-amound}
ggplot(train, aes(x = TransactionAmt, fill = factor(isFraud))) +
  geom_histogram(bins = 100, position = "identity", alpha = 0.6) +
  scale_x_log10(labels = scales::dollar) +
  scale_fill_manual(
    values = c("0" = "#3B8BD4", "1" = "#D85A30"),
    labels = c("0" = "Legitimate", "1" = "Fraud")
  ) +
  labs(
    title = "Transaction Amount Distribution (Log Scale)",
    x = "Transaction Amount (log scale)",
    y = "Count",
    fill = "Transaction type"
  ) +
  theme_minimal()
```

**Observation**:

- The majority of transactions cluster between $10 and $1,000, with a
  peak around $50–$100 suggesting common purchase amounts.
- Fraudulent transactions follow a distribution very similar to legitimate
  ones in terms of amount — `TransactionAmt` alone is not a strong
  discriminator between classes.
- The spikes at round numbers (e.g., $100, $200) suggest transaction
  rounding behavior, which may be informative as an engineered binary
  feature (e.g., `is_round_amount`).
- Log scaling is necessary to visualize the distribution — the raw
  amounts are heavily right-skewed, which justifies a `log(TransactionAmt)`
  transformation in the feature engineering pipeline.

### 5.2 Device Info

`DeviceInfo` and `DeviceType` come from `train_identity`, which covers
only 144,233 of 590,540 transactions (~24%). The remaining ~76% of
missing values after the join are structural — not a data quality issue.

```{r device-info-distribution}
# Distribution within the identity subset (train_identity directly)
train_identity |>
  filter(!is.na(DeviceInfo), DeviceInfo != "") |>
  count(DeviceInfo) |>
  slice_max(n, n = 10) |>
  ggplot(aes(x = reorder(DeviceInfo, n), y = n)) +
  geom_col(fill = "#3B8BD4") +
  coord_flip() +
  labs(
    title = "Top 10 Device Info (identity subset only)",
    x = "Device Info", y = "Count"
  ) +
  theme_minimal()
```

```{r device-type-fraud-rate}
# Fraud rate by DeviceType — requires join for isFraud
train |>
  mutate(DeviceType = replace_na(DeviceType, "missing")) |>
  count(DeviceType, isFraud) |>
  group_by(DeviceType) |>
  mutate(fraud_rate = n / sum(n)) |>
  filter(isFraud == 1) |>
  ggplot(aes(x = reorder(DeviceType, fraud_rate), y = fraud_rate)) +
  geom_col(fill = "#D85A30") +
  coord_flip() +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "Fraud Rate by Device Type",
    x = "Device Type", y = "Fraud Rate"
  ) +
  theme_minimal()
```

```{r os-family-fraud-rate}
# OS family grouping — requires join for isFraud
# DeviceInfo contains truncated UA strings, not full user agents.
# Build/ pattern identifies Android model numbers (e.g. SM-G930V Build/NRD90M)
train |>
  filter(!is.na(DeviceInfo), DeviceInfo != "") |>
  mutate(os_family = case_when(
    str_detect(DeviceInfo, regex("windows|trident|rv:", ignore_case = TRUE)) ~ "Windows/IE",
    str_detect(DeviceInfo, regex("ios device|iphone|ipad", ignore_case = TRUE)) ~ "iOS",
    str_detect(DeviceInfo, regex("macos|mac os", ignore_case = TRUE)) ~ "MacOS",
    str_detect(DeviceInfo, regex("android|Build/|SM-|samsung|redmi|lenovo|lg-|xt[0-9]",
      ignore_case = TRUE
    )) ~ "Android",
    TRUE ~ "Other"
  )) |>
  count(os_family, isFraud) |>
  group_by(os_family) |>
  mutate(total = sum(n), fraud_rate = n / total) |>
  filter(isFraud == 1, total >= 100) |>
  ggplot(aes(x = reorder(os_family, fraud_rate), y = fraud_rate)) +
  geom_col(fill = "#D85A30") +
  coord_flip() +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "Fraud Rate by OS Family",
    x = "OS Family", y = "Fraud Rate"
  ) +
  theme_minimal()
```

**Observations:**

- The ~76% missingness in `DeviceInfo` after the join is structural —
  only ~24% of transactions have identity records. Within that subset,
  missingness drops to 18% for `DeviceInfo` and 2% for `DeviceType`.
- Windows dominates the identity subset (~48,000 records), followed
  by iOS (~20,000) and MacOS (~13,000). `Trident/7.0` and `rv:11.0`
  are IE11 user agent fragments — grouped with Windows/IE.
- `DeviceInfo` contains truncated user agent strings, not full UA
  strings — standard UA parsing libraries are unreliable on this data.
  OS family grouping uses pattern matching: `Build/` identifies Android
  model numbers. The `Other` category warrants further investigation
  in feature engineering.
- **Device type**: mobile shows notably higher fraud rate (>10%) vs
  desktop (~6.25%). Missing device type has very low fraud rate
  (<2.5%) — consistent with lower-risk transaction channels where
  device fingerprinting is unavailable.
- **OS family**: Android has the highest fraud rate (~13%),
  significantly above iOS (~6%), Windows/IE (~6%), Other (~5%),
  and MacOS (~2.5%).
- The `missing` category in `DeviceType` (~2.5% fraud rate) aggregates
  all NAs including the ~446,307 structural missing values. Its low
  fraud rate reflects dilution by the large volume of transactions
  without identity records — not that missing device info predicts
  lower fraud.
- `Other` in `os_family` excludes NAs entirely (filtered before
  grouping) — it represents only transactions with a `DeviceInfo`
  value that did not match any OS pattern. These are likely obscure
  device strings worth investigating in feature engineering.
- **`has_identity`** is therefore critical: it separates the structural
  missing (~76%) from the actual device populations, preventing the
  dilution effect seen in the `missing` DeviceType category.

- **Pipeline implications**:
  - `has_identity`: binary feature encoding absence of identity record
  - `os_family`: OS grouped from `DeviceInfo` using `Build/` pattern
  - `DeviceType`: label encoded (mobile/desktop/missing)
  - `Other` in `os_family` investigated further in feature engineering

### 5.3 Temporal Features Analysis

`TransactionDT` is a timedelta in seconds from an undisclosed reference
point. The absolute origin is arbitrary — we use "2017-12-01" solely to
convert seconds into interpretable hour and day-of-week values. The
competition data is from 2019, but absolute dates are not recoverable
from this dataset. Relative temporal patterns (hour of day, day of week)
are valid regardless of the chosen origin.

```{r temporal-features}
# Convert TransactionDT to hours and days
# Note: the dataset starts at a reference point, we define a fake 'origin'
origin_time <- as.POSIXct("2017-12-01", tz = "UTC")

train <- train |>
  mutate(
    TransactionDate = origin_time + TransactionDT,
    TransactionHour = lubridate::hour(TransactionDate),
    TransactionDay  = lubridate::wday(TransactionDate, label = TRUE)
  )
```

next step is visualizing the distribution of transactions by hour and day of the week. This helps us understand if there are specific times or days when fraudulent transactions are more likely to occur.

```{r temporal-features-plot}
# Plot fraud rate by hour
ggplot(train, aes(x = TransactionHour, fill = factor(isFraud))) +
  geom_histogram(binwidth = 1, position = "fill") +
  labs(
    title = "Fraud Ratio by Hour of Day",
    x = "Hour of Day",
    y = "Proportion",
    fill = "Is Fraud"
  ) +
  scale_fill_manual(
    values = c("0" = "#3B8BD4", "1" = "#D85A30"),
    labels = c("0" = "Legitimate", "1" = "Fraud")
  ) +
  scale_x_continuous(breaks = 0:23) +
  theme_minimal()
```

**Observation:**

This plot shows the proportion of fraudulent vs. non-fraudulent transactions across different hours of the day.

**Hour of day:**
- Most transactions occur between 5 a.m. and 12 p.m. (UTC reference).
- Fraudulent activity peaks between 6–9 a.m., with a maximum around
  7 a.m. — consistent with automated fraud attempts during low
  supervision periods or bot activity scheduled around timezone patterns.
- **Pipeline implication**: `TransactionHour` will be included as a
  cyclical feature using sin/cos encoding to preserve the circular
  nature of time (hour 23 is close to hour 0).

This information could be useful to:

- Engineer features like `IsMorning`, `IsNight`, etc.
- Apply time-aware models or rules.

```{r temporal-features-plot2}
ggplot(train, aes(x = TransactionDay, fill = factor(isFraud))) +
  geom_bar(position = "fill") + # bars with separation for readability
  labs(
    title = "Fraud Ratio by Day of Week",
    x = "Day of Week",
    y = "Proportion",
    fill = "Is Fraud"
  ) +
  scale_fill_manual(
    values = c("0" = "#3B8BD4", "1" = "#D85A30"),
    labels = c("0" = "Legitimate", "1" = "Fraud")
  ) +
  theme_minimal()
```

**Observation:**

This bar plot shows the proportion of fraudulent transactions by day of the week.

**Day of week:**
- Fraud rate is remarkably stable across all days — no significant
  peaks or drops.
- This uniformity suggests automated fraud systems operating
  independently of calendar patterns.
- `TransactionDay` has limited standalone predictive power but may
  contribute signal in interaction features (e.g., `hour × day`).
- **Pipeline implication**: included as a low-priority feature —
  interaction with `TransactionHour` more valuable than raw day alone.

As a result, `TransactionDay` may have **limited predictive power on its own**, but could still be useful when combined with other temporal or categorical features.
