---
title: "Exploratory Data Analysis (EDA) - IEEE-CIS Fraud Detection"
author: "Manuel Alejandro Matías Astorga"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: false
    theme: flatly
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 10,
  fig.height = 6
)
```

## 1. Introduction

This notebook presents an Exploratory Data Analysis (EDA) of the [IEEE-CIS Fraud Detection](https://www.kaggle.com/competitions/ieee-fraud-detection) dataset, provided by Vesta Corporation for a Kaggle competition. The dataset contains anonymized transactional and identity information with the the goal of detecting fraudulent transactions.

**Key questions this EDA answers:**

- How severe is the class imbalance, and what does it imply for modeling?
- What patterns in transaction amount, time, and device distinguish fraud from legitimate transactions?
- Which features carry the most signal for fraud detection?
- What multicollinearity exists among predictors?
- Where are the missing values concentrated, and are they informative?

## 2. Libraries

We'll require the next libraries for data manipulation and visualization.

```{r libraries}
library(tidyverse)
library(arrow) # read parquet — replaces data.table for I/O
library(gridExtra)
library(skimr)
library(ggcorrplot)
library(naniar)
```

## 3. Data Loading

The dataset is composed of two main files:

- **`train_transaction`**: transactional data for online purchases — amount,
  product codes, card and address information, and the fraud label (`isFraud`).

- **`train_identity`**: identity-related features (device info, browser, email
  domain) for a subset of transactions. Only ~24% of transactions have
  corresponding identity information.

To perform a comprehensive analysis, both datasets are merged using
`TransactionID` as the join key. A **left join** preserves all transactions
from `train_transaction`, including those without identity information.
This ensures:

- The **full set of transactions** is available for fraud analysis.
- Identity features are included **only when available** — their absence
  is itself informative and will be considered during feature engineering.

```{r data-loading}
train_transaction <- read_parquet(
  "../../data/processed/train_transaction.parquet"
)
train_identity <- read_parquet(
  "../../data/processed/train_identity.parquet"
)

cat(
  "train_transaction:", nrow(train_transaction), "rows,",
  ncol(train_transaction), "cols\n"
)
cat(
  "train_identity:   ", nrow(train_identity), "rows,",
  ncol(train_identity), "cols\n"
)
```

```{r data-join}
train <- train_transaction |>
  left_join(train_identity, by = "TransactionID")

cat("Combined dataset:", nrow(train), "rows,", ncol(train), "cols\n")
```

`train_transaction` contains 590,540 rows and 394 columns, while
`train_identity` has 144,233 rows and 41 columns. The merged dataset
has **590,540 rows and 434 columns** — `TransactionID` appears in both
files but is not duplicated after the join.

Due to the high dimensionality and anonymized nature of the dataset
(434 features), this EDA focuses on interpretable and structurally
relevant features: `TransactionAmt`, `DeviceType`, email domains,
`C1–C14`, and `D1–D15`. Feature importance and selection will be
addressed in the modeling stage using SHAP values.
