---
title: "Exploratory Data Analysis (EDA) - IEEE-CIS Fraud Detection"
author: "Manuel Alejandro Matías Astorga"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: false
    theme: flatly
    highlight: tango
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 10,
  fig.height = 6
)
```

## 1. Introduction

This notebook presents an Exploratory Data Analysis (EDA) of the
[IEEE-CIS Fraud
Detection](https://www.kaggle.com/competitions/ieee-fraud-detection)
dataset, provided by Vesta Corporation for a Kaggle competition. The
dataset contains anonymized transactional and identity information with
the the goal of detecting fraudulent transactions.

**Key questions this EDA answers:**

- How severe is the class imbalance, and what does it imply for
  modeling?
- What patterns in transaction amount, time, and device distinguish
  fraud from legitimate transactions?
- Which features carry the most signal for fraud detection?
- What multicollinearity exists among predictors?
- Where are the missing values concentrated, and are they informative?

## 2. Libraries

We'll require the next libraries for data manipulation and
visualization.

```{r libraries}
library(tidyverse)
library(arrow)
library(gridExtra)
library(skimr)
library(ggcorrplot)
library(naniar)
library(here)
```

## 3. Data Loading

The dataset is composed of two main files:

- **`train_transaction`**: transactional data for online purchases —
  amount, product codes, card and address information, and the fraud
  label (`isFraud`).

- **`train_identity`**: identity-related features (device info, browser,
  email domain) for a subset of transactions. Only \~24% of transactions
  have corresponding identity information.

To perform a comprehensive analysis, both datasets are merged using
`TransactionID` as the join key. A **left join** preserves all
transactions from `train_transaction`, including those without identity
information. This ensures:

- The **full set of transactions** is available for fraud analysis.
- Identity features are included **only when available** — their absence
  is itself informative and will be considered during feature
  engineering.

```{r data-loading}
train_transaction <- read_parquet(
  "../../data/processed/train_transaction.parquet"
)
train_identity <- read_parquet(
  "../../data/processed/train_identity.parquet"
)

cat(
  "train_transaction:", nrow(train_transaction), "rows,",
  ncol(train_transaction), "cols\n"
)
cat(
  "train_identity:   ", nrow(train_identity), "rows,",
  ncol(train_identity), "cols\n"
)
```

```{r data-join}
train <- train_transaction |>
  left_join(train_identity, by = "TransactionID")

cat("Combined dataset:", nrow(train), "rows,", ncol(train), "cols\n")
```

`train_transaction` contains 590,540 rows and 394 columns, while
`train_identity` has 144,233 rows and 41 columns. The merged dataset has
**590,540 rows and 434 columns** — `TransactionID` appears in both files
but is not duplicated after the join.

Due to the high dimensionality and anonymized nature of the dataset (434
features), this EDA focuses on interpretable and structurally relevant
features: `TransactionAmt`, `DeviceType`, email domains, `C1–C14`, and
`D1–D15`. Feature importance and selection will be addressed in the
modeling stage using SHAP values.

## 4. Data Overview

Let's analyze the dataset to understand its structure, missing values,
and class balance.

### 4.1 Missing Values

To assess the quality of the data, we calculated the percentage of
missing values in each column using the
`summarise_all(~ mean(is.na(.)))` function chain. This provides a quick
overview of how much information is missing in each feature.

```{r missing-values}
missing_data <- train |>
  summarise(across(everything(), ~ mean(is.na(.)))) |>
  pivot_longer(
    cols = everything(),
    names_to = "variable",
    values_to = "missing_perc"
  ) |>
  arrange(desc(missing_perc))

# View top 20 variables with the highest percentage of missing values
head(missing_data, 20)
```

From the results above, we observe that several variables (especially
`id_24`, `id_25`, `id_07`, `id_08`, and many features from the `id_` and
`D` families) have over 85–99% missing values. This is critical because:

- Such high missingness may suggest that these features are either
  optional, sparsely recorded, or system-dependent.
- These features may not be useful for modeling without appropriate
  imputation, encoding, or dimensionality reduction.
- We must decide whether to drop these variables or retain them by
  filling missing values (e.g., using domain knowledge or imputation
  strategies).

This information will be key during the feature selection and
preprocessing phases.

### 4.2 Class Balance

This dataset is highly imbalanced, with a strong predominance of
non-fraudulent transactions. Understanding this imbalance is crucial
before training machine learning models, as it can bias the models
toward the majority class.

```{r class-balance}
# Distribution of target variable
train |>
  count(isFraud) |>
  mutate(percentage = n / sum(n))
```

**Observation**:

The target variable `isFraud` is heavily imbalanced:

- **Non-fraudulent transactions** (isFraud = 0): 569,877 (\~96.5%)
- **Fraudulent transactions** (isFraud = 1): 20,663 (\~3.5%)

This imbalance has direct implications for modeling and evaluation:

- **Accuracy is misleading**: a model predicting "no fraud" for every
  transaction achieves 96.5% accuracy while detecting zero fraud cases.
  AUC-ROC and Average Precision (AUC-PR) will be used instead.

- **Class weighting over resampling**: `scale_pos_weight` in LightGBM
  penalizes misclassification of the minority class without generating
  synthetic data. Resampling techniques like SMOTE are avoided —
  generating synthetic fraudulent transactions in a 400+ dimensional
  space produces samples that do not reflect real fraud patterns.

- **Threshold tuning**: the default classification threshold of 0.5 is
  suboptimal for imbalanced datasets. Since this is a Kaggle competition
  evaluated on AUC-ROC — a threshold-free metric — threshold selection
  is not critical here, but would be a key decision in a production
  deployment where false positive and false negative costs differ.

## 5. Exploratory Visualizations

### 5.1 Transaction Amount Distribution

```{r transaction-amount}
trans_amount <- ggplot(train, aes(x = TransactionAmt, fill = factor(isFraud))) +
  geom_histogram(bins = 100, position = "identity", alpha = 0.6) +
  scale_x_log10(labels = scales::dollar) +
  scale_fill_manual(
    values = c("0" = "#3B8BD4", "1" = "#D85A30"),
    labels = c("0" = "Legitimate", "1" = "Fraud")
  ) +
  labs(
    title = "Transaction Amount Distribution (Log Scale)",
    x = "Transaction Amount (log scale)",
    y = "Count",
    fill = "Transaction type"
  ) +
  theme_minimal()
trans_amount
ggsave(
  filename = here("reports", "figures", "transaction_amt_dist.png"),
  plot = trans_amount,
  width = 10,
  height = 6,
  dpi = "print"
)
```

**Observation**:

- The majority of transactions cluster between \$10 and \$1,000, with a
  peak around \$50–\$100 suggesting common purchase amounts.
- Fraudulent transactions follow a distribution very similar to
  legitimate ones in terms of amount — `TransactionAmt` alone is not a
  strong discriminator between classes.
- The spikes at round numbers (e.g., \$100, \$200) suggest transaction
  rounding behavior, which may be informative as an engineered binary
  feature (e.g., `is_round_amount`).
- Log scaling is necessary to visualize the distribution — the raw
  amounts are heavily right-skewed, which justifies a
  `log(TransactionAmt)` transformation in the feature engineering
  pipeline.

### 5.2 Device Info

`DeviceInfo` and `DeviceType` come from `train_identity`, which covers
only 144,233 of 590,540 transactions (\~24%). The remaining \~76% of
missing values after the join are structural — not a data quality issue.

```{r device-info-distribution}
# Distribution within the identity subset (train_identity directly)
train_identity |>
  filter(!is.na(DeviceInfo), DeviceInfo != "") |>
  count(DeviceInfo) |>
  slice_max(n, n = 10) |>
  ggplot(aes(x = reorder(DeviceInfo, n), y = n)) +
  geom_col(fill = "#3B8BD4") +
  coord_flip() +
  labs(
    title = "Top 10 Device Info (identity subset only)",
    x = "Device Info", y = "Count"
  ) +
  theme_minimal()
```

```{r device-type-fraud-rate}
# Fraud rate by DeviceType — requires join for isFraud
train |>
  mutate(DeviceType = replace_na(DeviceType, "missing")) |>
  count(DeviceType, isFraud) |>
  group_by(DeviceType) |>
  mutate(fraud_rate = n / sum(n)) |>
  filter(isFraud == 1) |>
  ggplot(aes(x = reorder(DeviceType, fraud_rate), y = fraud_rate)) +
  geom_col(fill = "#D85A30") +
  coord_flip() +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "Fraud Rate by Device Type",
    x = "Device Type", y = "Fraud Rate"
  ) +
  theme_minimal()
```

```{r os-family-fraud-rate}
# OS family grouping — requires join for isFraud
# DeviceInfo contains truncated UA strings, not full user agents.
# Build/ pattern identifies Android model numbers (e.g. SM-G930V Build/NRD90M)
train |>
  filter(!is.na(DeviceInfo), DeviceInfo != "") |>
  mutate(os_family = case_when(
    str_detect(DeviceInfo, regex("windows|trident|rv:", ignore_case = TRUE)) ~ "Windows/IE",
    str_detect(DeviceInfo, regex("ios device|iphone|ipad", ignore_case = TRUE)) ~ "iOS",
    str_detect(DeviceInfo, regex("macos|mac os", ignore_case = TRUE)) ~ "MacOS",
    str_detect(DeviceInfo, regex("android|Build/|SM-|samsung|redmi|lenovo|lg-|xt[0-9]",
      ignore_case = TRUE
    )) ~ "Android",
    TRUE ~ "Other"
  )) |>
  count(os_family, isFraud) |>
  group_by(os_family) |>
  mutate(total = sum(n), fraud_rate = n / total) |>
  filter(isFraud == 1, total >= 100) |>
  ggplot(aes(x = reorder(os_family, fraud_rate), y = fraud_rate)) +
  geom_col(fill = "#D85A30") +
  coord_flip() +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "Fraud Rate by OS Family",
    x = "OS Family", y = "Fraud Rate"
  ) +
  theme_minimal()
```

**Observations:**

- The \~76% missingness in `DeviceInfo` after the join is structural —
  only \~24% of transactions have identity records. Within that subset,
  missingness drops to 18% for `DeviceInfo` and 2% for `DeviceType`.

- Windows dominates the identity subset (\~48,000 records), followed by
  iOS (\~20,000) and MacOS (\~13,000). `Trident/7.0` and `rv:11.0` are
  IE11 user agent fragments — grouped with Windows/IE.

- `DeviceInfo` contains truncated user agent strings, not full UA
  strings — standard UA parsing libraries are unreliable on this data.
  OS family grouping uses pattern matching: `Build/` identifies Android
  model numbers. The `Other` category warrants further investigation in
  feature engineering.

- **Device type**: mobile shows notably higher fraud rate (\>10%) vs
  desktop (\~6.25%). Missing device type has very low fraud rate
  (\<2.5%) — consistent with lower-risk transaction channels where
  device fingerprinting is unavailable.

- **OS family**: Android has the highest fraud rate (\~13%),
  significantly above iOS (\~6%), Windows/IE (\~6%), Other (\~5%), and
  MacOS (\~2.5%).

- The `missing` category in `DeviceType` (\~2.5% fraud rate) aggregates
  all NAs including the \~446,307 structural missing values. Its low
  fraud rate reflects dilution by the large volume of transactions
  without identity records — not that missing device info predicts lower
  fraud.

- `Other` in `os_family` excludes NAs entirely (filtered before
  grouping) — it represents only transactions with a `DeviceInfo` value
  that did not match any OS pattern. These are likely obscure device
  strings worth investigating in feature engineering.

- **`has_identity`** is therefore critical: it separates the structural
  missing (\~76%) from the actual device populations, preventing the
  dilution effect seen in the `missing` DeviceType category.

- **Pipeline implications**:

  - `has_identity`: binary feature encoding absence of identity record
  - `os_family`: OS grouped from `DeviceInfo` using `Build/` pattern
  - `DeviceType`: label encoded (mobile/desktop/missing)
  - `Other` in `os_family` investigated further in feature engineering

### 5.3 Temporal Features Analysis

`TransactionDT` is a timedelta in seconds from an undisclosed reference
point. The absolute origin is arbitrary — we use "2017-12-01" solely to
convert seconds into interpretable hour and day-of-week values. The
competition data is from 2019, but absolute dates are not recoverable
from this dataset. Relative temporal patterns (hour of day, day of week)
are valid regardless of the chosen origin.

```{r temporal-features}
# Convert TransactionDT to hours and days
# Note: the dataset starts at a reference point, we define a fake 'origin'
origin_time <- as.POSIXct("2017-12-01", tz = "UTC")

train <- train |>
  mutate(
    TransactionDate = origin_time + TransactionDT,
    TransactionHour = lubridate::hour(TransactionDate),
    TransactionDay  = lubridate::wday(TransactionDate, label = TRUE)
  )
```

next step is visualizing the distribution of transactions by hour and
day of the week. This helps us understand if there are specific times or
days when fraudulent transactions are more likely to occur.

```{r temporal-features-plot}
# Plot fraud rate by hour
ggplot(train, aes(x = TransactionHour, fill = factor(isFraud))) +
  geom_histogram(binwidth = 1, position = "fill") +
  labs(
    title = "Fraud Ratio by Hour of Day",
    x = "Hour of Day",
    y = "Proportion",
    fill = "Is Fraud"
  ) +
  scale_fill_manual(
    values = c("0" = "#3B8BD4", "1" = "#D85A30"),
    labels = c("0" = "Legitimate", "1" = "Fraud")
  ) +
  scale_x_continuous(breaks = 0:23) +
  theme_minimal()
```

**Observation:**

This plot shows the proportion of fraudulent vs. non-fraudulent
transactions across different hours of the day.

**Hour of day:** - Most transactions occur between 5 a.m. and 12 p.m.
(UTC reference). - Fraudulent activity peaks between 6–9 a.m., with a
maximum around 7 a.m. — consistent with automated fraud attempts during
low supervision periods or bot activity scheduled around timezone
patterns. - **Pipeline implication**: `TransactionHour` will be included
as a cyclical feature using sin/cos encoding to preserve the circular
nature of time (hour 23 is close to hour 0).

This information could be useful to:

- Engineer features like `IsMorning`, `IsNight`, etc.
- Apply time-aware models or rules.

```{r temporal-features-plot2}
ggplot(train, aes(x = TransactionDay, fill = factor(isFraud))) +
  geom_bar(position = "fill") + # bars with separation for readability
  labs(
    title = "Fraud Ratio by Day of Week",
    x = "Day of Week",
    y = "Proportion",
    fill = "Is Fraud"
  ) +
  scale_fill_manual(
    values = c("0" = "#3B8BD4", "1" = "#D85A30"),
    labels = c("0" = "Legitimate", "1" = "Fraud")
  ) +
  theme_minimal()
```

**Observation:**

This bar plot shows the proportion of fraudulent transactions by day of
the week.

**Day of week:** - Fraud rate is remarkably stable across all days — no
significant peaks or drops. - This uniformity suggests automated fraud
systems operating independently of calendar patterns. - `TransactionDay`
has limited standalone predictive power but may contribute signal in
interaction features (e.g., `hour × day`). - **Pipeline implication**:
included as a low-priority feature — interaction with `TransactionHour`
more valuable than raw day alone.

As a result, `TransactionDay` may have **limited predictive power on its
own**, but could still be useful when combined with other temporal or
categorical features.

### 5.4 Categorical Features Analysis

We will explore some categorical features that may provide insights into
the fraud detection process. The following features are of particular
interest:

- `ProductCD`: Product code associated with the transaction.
- `card4`: Card type (e.g., Visa, Mastercard).
- `card6`: Card type (e.g., credit, debit).
- `P_emaildomain`: Email domain of the payer.
- `R_emaildomain`: Email domain of the recipient.

We analyze the fraud proportions across these categories using
proportion-based bar plots (`position = "fill"` in `ggplot2`) to better
visualize class imbalance across each level.

- `ProductCD`

```{r productcd-barplot}
ggplot(train, aes(x = ProductCD, fill = factor(isFraud))) +
  geom_bar(position = "fill") +
  labs(
    title = "Fraud Proportion by ProductCD",
    x = "ProductCD",
    y = "Proportion",
    fill = "Is Fraud"
  ) +
  scale_fill_manual(
    values = c("0" = "#3B8BD4", "1" = "#D85A30"),
    labels = c("0" = "Legitimate", "1" = "Fraud")
  ) +
  theme_minimal()
```

**Observation**:

- Product code C has a noticeably higher fraud rate compared to the
  others.

- This may suggest that some product categories are more prone to fraud,
  perhaps due to their price range, popularity, or nature of
  transactions.

- `card4`

```{r card4-barplot}
ggplot(train, aes(x = card4, fill = factor(isFraud))) +
  geom_bar(position = "fill") +
  labs(
    title = "Fraud Proportion by Card Type (card4)",
    x = "Card Type",
    y = "Proportion",
    fill = "Is Fraud"
  ) +
  scale_fill_manual(
    values = c("0" = "#3B8BD4", "1" = "#D85A30"),
    labels = c("0" = "Legitimate", "1" = "Fraud")
  ) +
  theme_minimal()
```

**Observation**:

- Among card types, `discover` shows the highest fraud proportion.

- `Visa` and `Mastercard` dominate in volume, but their fraud ratio is
  relatively lower.

- `card6`

```{r card6-barplot}
ggplot(train, aes(x = card6, fill = factor(isFraud))) +
  geom_bar(position = "fill") +
  labs(
    title = "Fraud Proportion by Card Type (card6)",
    x = "Card Type (card6)",
    y = "Proportion",
    fill = "Is Fraud"
  ) +
  scale_fill_manual(
    values = c("0" = "#3B8BD4", "1" = "#D85A30"),
    labels = c("0" = "Legitimate", "1" = "Fraud")
  ) +
  theme_minimal()
```

**Observation**:

- Fraud is more frequent in `credit` cards compared to `debit` or
  `charge` cards.

- Transactions labeled as `debit or credit` are rare and also show lower
  fraud prevalence, this may due to the duplication of options since
  there is a separate `credit` and `debit` option..

- `P_emaildomain`

```{r p-emaildomain-barplot}
# Top 10 most common purchaser email domains
top_p_domains <- train |>
  count(P_emaildomain) |>
  slice_max(n, n = 10) |>
  pull(P_emaildomain)

ggplot(
  train |> filter(P_emaildomain %in% top_p_domains),
  aes(x = P_emaildomain, fill = factor(isFraud))
) +
  geom_bar(position = "fill") +
  labs(
    title = "Fraud Proportion by Purchaser Email Domain",
    x = "Purchaser Email Domain (Top 10)",
    y = "Proportion",
    fill = "Is Fraud"
  ) +
  scale_fill_manual(
    values = c("0" = "#3B8BD4", "1" = "#D85A30"),
    labels = c("0" = "Legitimate", "1" = "Fraud")
  ) +
  theme_minimal() +
  coord_flip()
```

```{r p-emaildomain-fraud-rate }
p_email_domain_fr <- train |>
  filter(!is.na(P_emaildomain)) |>
  count(P_emaildomain, isFraud) |>
  group_by(P_emaildomain) |>
  mutate(total = sum(n), fraud_rate = n / total) |>
  ungroup() |>
  filter(isFraud == 1, total >= 500) |>
  slice_max(fraud_rate, n = 10)


ggplot(p_email_domain_fr, aes(x = reorder(P_emaildomain, fraud_rate), y = fraud_rate)) +
  geom_col(fill = "#D85A30") +
  coord_flip() +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "Top 10 Purchaser Email Domains by Fraud Rate (min. 500 transactions)",
    x = "Purchaser Email Domain",
    y = "Fraud Rate"
  ) +
  theme_minimal()
```

**Observation**:

- Among the top 10 purchaser domains by volume, `outlook.com` shows the
  highest fraud proportion (\~12.5%), followed by `hotmail.com` (\~6%)
  and `gmail.com` (\~5.5%). High-volume providers like `gmail.com` and
  `yahoo.com` have relatively lower fraud rates despite their volume.

- The fraud rate plot (min. 500 transactions) reveals `mail.com` as the
  highest-risk purchaser domain (\~20%), followed by `outlook.com`
  (\~10%) and `live.com.mx` (\~5.5%). Microsoft-affiliated domains
  (`outlook`, `hotmail`, `live`) appear consistently across the top —
  possibly reflecting account compromise patterns on older email
  platforms.

- **Pipeline implication**: `P_emaildomain` will be frequency-encoded.
  Microsoft-affiliated domains (`outlook.com`, `hotmail.com`,
  `live.com`, `live.com.mx`) may warrant grouping into a single
  category.

- `R_emaildomain`

```{r r-emaildomain-barplot}
# Top 10 most common recipient email domains
top_r_domains <- train |>
  count(R_emaildomain) |>
  slice_max(n, n = 10) |>
  pull(R_emaildomain)

ggplot(
  train |> filter(R_emaildomain %in% top_r_domains),
  aes(x = R_emaildomain, fill = factor(isFraud))
) +
  geom_bar(position = "fill") +
  labs(
    title = "Fraud Proportion by Recipient Email Domain",
    x = "Recipient Email Domain (Top 10)",
    y = "Proportion",
    fill = "Is Fraud"
  ) +
  scale_fill_manual(
    values = c("0" = "#3B8BD4", "1" = "#D85A30"),
    labels = c("0" = "Legitimate", "1" = "Fraud")
  ) +
  theme_minimal() +
  coord_flip()
```

```{r r-emaildomain-fraud-rate }
r_email_domain_fr <- train |>
  filter(!is.na(R_emaildomain)) |>
  count(R_emaildomain, isFraud) |>
  group_by(R_emaildomain) |>
  mutate(total = sum(n), fraud_rate = n / total) |>
  ungroup() |>
  filter(isFraud == 1, total >= 500) |>
  slice_max(fraud_rate, n = 10)


ggplot(r_email_domain_fr, aes(x = reorder(R_emaildomain, fraud_rate), y = fraud_rate)) +
  geom_col(fill = "#D85A30") +
  coord_flip() +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "Top 10 Recipient Email Domains by Fraud Rate (min. 500 transactions)",
    x = "Recipient Email Domain",
    y = "Fraud Rate"
  ) +
  theme_minimal()
```

**Observation:**

- Recipient domains show notably higher fraud rates than purchaser
  domains for the same providers — `outlook.com` leads at \~17.5%,
  followed by `icloud.com` (\~12.5%), `gmail.com` (\~12%), and
  `hotmail.com` (\~8-10%).
- `anonymous.com` appears in the recipient top 10 (\~2.5%) despite lower
  volume — its presence as a recipient domain is a strong fraud signal
  given its name suggests intentional anonymization.
- The asymmetry between purchaser and recipient fraud rates is notable:
  a transaction where the recipient uses `outlook.com` is riskier than
  one where the purchaser uses the same domain. This suggests recipient
  email domain carries more signal than purchaser email domain.
- **Pipeline implication**: `R_emaildomain` will be frequency-encoded. A
  binary feature `p_r_domain_match` (purchaser domain == recipient
  domain) captures whether both sides of the transaction share the same
  email provider — an unusual pattern worth investigating.
- `anonymous.com` appears as a recipient domain with \~2.5% fraud rate —
  below the dataset average of \~3.5%. Its presence is notable given the
  name implies intentional anonymization, but the data does not support
  flagging it as high-risk based on fraud rate alone. It warrants
  monitoring but not automatic flagging.

### 5.5 Continuous Features Analysis - C1 to C14

See [this
post](https://www.kaggle.com/c/ieee-fraud-detection/discussion/101203)
for more information about these features and other.

`C1–C14` are **counting features** — each represents a count of entities
associated with a payment card, such as number of addresses, devices, or
emails linked to that card. The actual meaning of each variable is
masked by Vesta, but their nature as counts has direct implications for
analysis:

- Count distributions are inherently right-skewed — most cards have few
  associated entities, while a small number have many.
- High counts may indicate account sharing (considering this is for a
  e-commerce payment solution, this meant it may be shared by kin or
  even friends), synthetic identities, or cards used across many
  transactions — all signals relevant to fraud.
- Filtering above the 99th percentile removes extreme outliers for
  visualization only — they are retained for modeling as they may encode
  the strongest fraud signals.

#### 5.5.1 C1 and C2 – Initial Exploration

C1 and C2 are the two continuous count variables with the widest range
and highest cardinality among the C features.

```{r c1-and-c2-distribution}
train |>
  select(C1, C2, isFraud) |>
  pivot_longer(cols = c(C1, C2), names_to = "Variable", values_to = "Value") |>
  ggplot(aes(x = Value, fill = factor(isFraud))) +
  geom_histogram(bins = 100, position = "identity", alpha = 0.5) +
  facet_wrap(~Variable, scales = "free") +
  scale_x_continuous(trans = "log1p") +
  scale_fill_manual(
    values = c("0" = "#3B8BD4", "1" = "#D85A30"),
    labels = c("0" = "Legitimate", "1" = "Fraud")
  ) +
  labs(
    title = "Distribution of C1 and C2 by Fraud Status",
    x = "Value (log scale)",
    y = "Count",
    fill = "Is Fraud"
  )
```

**Observation**:

- The distributions are highly skewed to the right.
- Fraudulent and non-fraudulent transactions overlap heavily in this
  space, though fine-grained differences might still be exploitable with
  modeling.
- A log transformation improves visualization but still shows
  long-tailed distributions.

To reduce noise from extreme outliers and improve interpretability, we
apply filtering below the 99th percentile.

```{r c1-and-c2-distribution-filtered}
# Calculating 99th percentiles for C1 and C2
c1_99 <- quantile(train$C1, 0.99, na.rm = TRUE)
c2_99 <- quantile(train$C2, 0.99, na.rm = TRUE)

# Filtering and plotting
train |>
  filter(C1 < c1_99, C2 < c2_99) |>
  select(C1, C2, isFraud) |>
  pivot_longer(cols = c(C1, C2), names_to = "Variable", values_to = "Value") |>
  ggplot(aes(x = Value, fill = factor(isFraud))) +
  geom_histogram(bins = 60, position = "identity", alpha = 0.5) +
  facet_wrap(~Variable, scales = "free") +
  scale_fill_manual(
    values = c("0" = "#3B8BD4", "1" = "#D85A30"),
    labels = c("0" = "Legitimate", "1" = "Fraud")
  ) +
  labs(
    title = "Distribution of C1 and C2 by Fraud Status (Filtered < 99th Percentile)",
    x = "Value",
    y = "Count",
    fill = "Is Fraud"
  )
```

**Note**: While we remove outliers for visualization clarity, they are
retained for modeling, as they may encode meaningful anomalies.

#### 5.5.2 General Analysis of C1–C14

Now, we extend the distribution analysis to the full set of C variables
grouped into two categories identified during exploration:
**continuous** (C1, C2, C5, C6, C9, C11, C13, C14) and **near-binary**
(C4, C7, C8, C10, C12). C3 is dropped — see Section 5.5.3 for the
rationale.

**Continuous count variables** have wide ranges and right-skewed
distributions with long tails extending to thousands.

```{r c-vars-plot-summary}
c_continuous <- c("C1", "C2", "C5", "C6", "C9", "C11", "C13", "C14")

c_long_cont <- train |>
  select(isFraud, all_of(c_continuous)) |>
  pivot_longer(cols = all_of(c_continuous), names_to = "Variable", values_to = "Value")

c_thresh_cont <- c_long_cont |>
  group_by(Variable) |>
  summarise(threshold = quantile(Value, 0.99, na.rm = TRUE), .groups = "drop")

c_long_cont <- c_long_cont |>
  left_join(c_thresh_cont, by = "Variable") |>
  filter(is.na(Value) | Value < threshold) |>
  mutate(Variable = factor(Variable, levels = c_continuous))

plot_c_block <- function(df, vars_subset) {
  df |>
    filter(Variable %in% vars_subset) |>
    mutate(Variable = factor(Variable, levels = vars_subset)) |>
    ggplot(aes(x = Value, fill = factor(isFraud))) +
    geom_histogram(bins = 60, position = "identity", alpha = 0.5) +
    facet_wrap(~Variable, scales = "free", ncol = 2) +
    scale_fill_manual(
      values = c("0" = "#3B8BD4", "1" = "#D85A30"),
      labels = c("0" = "Legitimate", "1" = "Fraud")
    ) +
    labs(x = "Value", y = "Count", fill = "Is Fraud") +
    theme_minimal()
}
```

```{r c-continuous-block-1}
plot_c_block(c_long_cont, c("C1", "C2", "C5", "C6"))
```

```{r c-continuous-block-2}
plot_c_block(c_long_cont, c("C9", "C11", "C13", "C14"))
```

**Near-binary count variables** are concentrated almost entirely at
values 0 or 1, with occasional values of 2–3 and extreme outliers above
the 99th percentile that are retained for modeling.

```{r c-vars-near-binary}
c_near_binary <- c("C4", "C7", "C8", "C10", "C12")

c_long_nb <- train |>
  select(isFraud, all_of(c_near_binary)) |>
  pivot_longer(cols = all_of(c_near_binary), names_to = "Variable", values_to = "Value")

c_thresh_nb <- c_long_nb |>
  group_by(Variable) |>
  summarise(threshold = quantile(Value, 0.99, na.rm = TRUE), .groups = "drop")

c_long_nb <- c_long_nb |>
  left_join(c_thresh_nb, by = "Variable") |>
  filter(is.na(Value) | Value < threshold) |>
  mutate(Variable = factor(Variable, levels = c_near_binary))

c_long_nb |>
  filter(Variable %in% c("C4", "C7", "C8", "C10", "C12")) |>
  mutate(Variable = factor(Variable, levels = c_near_binary)) |>
  ggplot(aes(x = Value, fill = factor(isFraud))) +
  geom_histogram(bins = 10, position = "identity", alpha = 0.5) +
  facet_wrap(~Variable, scales = "free", ncol = 3) +
  scale_fill_manual(
    values = c("0" = "#3B8BD4", "1" = "#D85A30"),
    labels = c("0" = "Legitimate", "1" = "Fraud")
  ) +
  labs(x = "Value", y = "Count", fill = "Is Fraud") +
  theme_minimal()
```

#### 5.5.3 C3 — Dropped Feature

C3 is dropped from analysis due to near-zero variation combined with
inverse fraud signal:

- **99.59%** of values are exactly 0.
- The remaining **2,429 records** (0.4%) have non-zero values (range
  1–26).
- Transactions where C3 \> 0 have a fraud rate of **0.21%** versus
  **3.51%** for C3 = 0 — the opposite of the expected direction.
- The feature has no actionable signal and would add noise to the model.

**Pipeline implication**: C3 is excluded from all downstream feature
engineering.

#### 5.5.4 Fraud Rate by Count Value

For count features, the most informative view is not the distribution
but the **fraud rate at each count level**. This reveals whether high
counts are associated with elevated fraud risk — the core question for
feature engineering.

**Near-binary variables (C4, C7, C8, C10, C12):** count values are
grouped as 0, 1, 2, and 3+ (categories with fewer than 50 records are
excluded).

```{r c-near-binary-fraud-rate}
train |>
  select(isFraud, all_of(c_near_binary)) |>
  pivot_longer(cols = all_of(c_near_binary), names_to = "Variable", values_to = "Value") |>
  filter(!is.na(Value)) |>
  mutate(Value = if_else(Value >= 3, "3+", as.character(as.integer(Value))),
         Value = factor(Value, levels = c("0", "1", "2", "3+"))) |>
  group_by(Variable, Value) |>
  summarise(n = n(), fraud_rate = mean(isFraud), .groups = "drop") |>
  filter(n >= 50) |>
  ggplot(aes(x = Value, y = fraud_rate)) +
  geom_col(fill = "#D85A30", width = 0.7) +
  geom_text(aes(label = scales::percent(fraud_rate, accuracy = 0.1)),
            vjust = -0.5, size = 3) +
  facet_wrap(~Variable, nrow = 1) +
  scale_y_continuous(labels = scales::percent, limits = c(0, 0.17)) +
  labs(
    title = "Fraud Rate by Count Value — Near-Binary C Variables",
    subtitle = "Categories with < 50 records excluded",
    x = "Count Value",
    y = "Fraud Rate"
  ) +
  theme_minimal()
```

**Key findings for near-binary variables:**

- Value = 2 consistently shows the highest fraud rate across C4, C7,
  C10, and C12 (10–14%), compared to 2–5% for value = 0.
- C8 is an exception: value = 1 shows the highest fraud rate (9.8%),
  suggesting its counting unit may differ from the others.
- All near-binary variables show monotonic fraud rate increase from
  value = 0 to value = 2 — clear discriminative signal.
- The jump from value = 1 to value = 2 represents a 3–6x fraud risk
  increase over baseline.

**Continuous count variables (C1, C2, C5, C6, C9, C11, C13, C14):**
fraud rate calculated per decile.

```{r c-continuous-fraud-rate}
train |>
  select(isFraud, all_of(c_continuous)) |>
  pivot_longer(cols = all_of(c_continuous), names_to = "Variable", values_to = "Value") |>
  filter(!is.na(Value)) |>
  group_by(Variable) |>
  mutate(decile = ntile(Value, 10)) |>
  ungroup() |>
  group_by(Variable, decile) |>
  summarise(n = n(), fraud_rate = mean(isFraud), .groups = "drop") |>
  ggplot(aes(x = decile, y = fraud_rate, fill = fraud_rate)) +
  geom_col() +
  facet_wrap(~Variable, scales = "free_y", ncol = 4) +
  scale_fill_gradient(low = "#f0c9be", high = "#D85A30", guide = "none") +
  scale_x_continuous(breaks = c(1, 5, 10), labels = c("D1\nlowest", "D5", "D10\nhighest")) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  labs(
    title = "Fraud Rate by Decile — Continuous C Variables",
    subtitle = "D1 = lowest 10% of counts, D10 = highest 10%",
    x = "Decile",
    y = "Fraud Rate"
  ) +
  theme_minimal()
```

**Key findings for continuous variables:**

- C1, C2, C6, C11, and C12 show the expected pattern: D10 (highest
  counts) associated with elevated fraud rates.
- **C13 and C14 exhibit an inverse pattern**: higher counts are
  associated with *lower* fraud rates. This may indicate that high
  counts of these entity types are associated with verified or
  legitimate accounts rather than fraud.
- **C9 shows a mid-range peak** at D2–D3 — unusual pattern worth
  monitoring.
- Extreme outliers (D10) are retained in modeling — they represent the
  strongest anomalies in count space.

### 5.5.5 Pipeline Implications — C Variables

| Feature / Group | Decision | Rationale |
|----|----|----|
| **C3** | Drop | 99.59% zeros; non-zero values have lower fraud rate than baseline — no predictive value |
| **C4, C7, C8, C10, C12** | Keep as-is | Near-binary distribution; clear fraud rate increase at value = 2; retained outliers are strongest signal |
| **C1, C2, C5, C6, C9, C11, C13, C14** | Keep as-is; note inverse patterns | Continuous with long tails; extreme outliers (D10) carry fraud signal; C13/C14 inverse direction worth monitoring in model |
| **All C variables** | Missing → -999 | LightGBM sentinel strategy (per architecture) — not explored here as all have 0% missing |
| **C13, C14** | Monitor in model | Inverse fraud-count relationship may indicate these count verified entities — SHAP will clarify importance |

### 5.6 Continuous Features Analysis - D1 to D15

The `D` variables are anonymous temporal or delay-related features.
Since their meaning is unknown, we group them into two sets based on
missingness:

| Set | Variables | Missing % | Non-missing n | Decision |
|----|----|----|----|----|
| **Complete** | D1, D2, D3, D4, D5, D10, D11, D15 | 0–52% | 280K–589K | Keep — full analysis |
| **High missingness** | D6, D7, D8, D9, D12, D13, D14 | 87–94% | 39K–75K | Drop — no actionable signal |

D6–D9 and D12–D14 have missingness of 87–94% with no structural
explanation (they are not explained by `has_identity`). The non-missing
observations do not show discriminative fraud patterns. These variables
are dropped from all further analysis. See Section 5.6.3 for the
complete pipeline decision table.

#### 5.6.1 Distribution of Complete D Variables

Distribution of the eight variables with lower missingness.

```{r d-vars-distribution}
d_complete <- c("D1", "D2", "D3", "D4", "D5", "D10", "D11", "D15")

d_long <- train |>
  select(isFraud, all_of(d_complete)) |>
  pivot_longer(cols = all_of(d_complete), names_to = "Variable", values_to = "Value") |>
  filter(!is.na(Value))

d_thresh <- d_long |>
  group_by(Variable) |>
  summarise(threshold = quantile(Value, 0.99, na.rm = TRUE), .groups = "drop")

d_long <- d_long |>
  left_join(d_thresh, by = "Variable") |>
  filter(Value < threshold) |>
  mutate(Variable = factor(Variable, levels = d_complete))

plot_d_block <- function(df, vars_subset) {
  df |>
    filter(Variable %in% vars_subset) |>
    mutate(Variable = factor(Variable, levels = vars_subset)) |>
    ggplot(aes(x = Value, fill = factor(isFraud))) +
    geom_histogram(bins = 50, position = "identity", alpha = 0.5) +
    facet_wrap(~Variable, scales = "free", ncol = 2) +
    scale_fill_manual(
      values = c("0" = "#3B8BD4", "1" = "#D85A30"),
      labels = c("0" = "Legitimate", "1" = "Fraud")
    ) +
    labs(x = "Value", y = "Count", fill = "Is Fraud") +
    theme_minimal()
}
```

```{r d-block-1}
plot_d_block(d_long, c("D1", "D2", "D3", "D4"))
```

```{r d-block-2}
plot_d_block(d_long, c("D5", "D10", "D11", "D15"))
```

**Observation:** All eight variables are right-skewed with heavy
concentration near low values and long tails. The overlap between fraud
and legitimate distributions makes visual distinction difficult — fraud
rate per decile analysis is more informative.

#### 5.6.2 Fraud Rate by Decile — Complete D Variables

Fraud rate per decile for the eight complete D variables.

```{r d-fraud-rate}
train |>
  select(isFraud, all_of(d_complete)) |>
  pivot_longer(cols = all_of(d_complete), names_to = "Variable", values_to = "Value") |>
  filter(!is.na(Value)) |>
  group_by(Variable) |>
  mutate(decile = ntile(Value, 10)) |>
  ungroup() |>
  group_by(Variable, decile) |>
  summarise(n = n(), fraud_rate = mean(isFraud), .groups = "drop") |>
  ggplot(aes(x = decile, y = fraud_rate, fill = fraud_rate)) +
  geom_col() +
  facet_wrap(~Variable, scales = "free_y", ncol = 4) +
  scale_fill_gradient(low = "#f0c9be", high = "#D85A30", guide = "none") +
  scale_x_continuous(breaks = c(1, 5, 10), labels = c("D1\nlowest", "D5", "D10\nhighest")) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  labs(
    title = "Fraud Rate by Decile — Complete D Variables",
    subtitle = "D1 = lowest 10% of values, D10 = highest 10%",
    x = "Decile",
    y = "Fraud Rate"
  ) +
  theme_minimal()
```

**Key findings:**

- **All eight variables show a clear inverse monotonic pattern** — lower
  values are associated with higher fraud rates. D1 has a fraud rate of
  4.6% while D10 has 1.2% (ratio 0.25). D2 shows the strongest inverse:
  D1 = 8.1% vs D10 = 0.8% (ratio 0.10).
- This is counter-intuitive for delay features: shorter delays should
  theoretically indicate legitimate, fast transactions, but the opposite
  holds. Transactions with minimal delay show the highest fraud rates.
- The inverse pattern may reflect that fraudsters operate automated
  systems that execute transactions very quickly — avoiding delays
  associated with manual verification or user interaction time.
- **D4, D7, D8, D14** show the widest fraud rate range (13–20%) across
  deciles — these may be the most discriminative D variables for the
  model despite low missingness.

#### 5.6.3 Pipeline Implications — D Variables

| Feature / Group | Decision | Rationale |
|----|----|----|
| **D6, D7, D8, D9** | Drop | 87–94% missing; no structural explanation; non-missing observations show no clear discriminative pattern |
| **D12, D13, D14** | Drop | 89% missing; no structural explanation; non-missing shows flat or mid-range peak patterns |
| **D1, D2, D3, D4, D5, D10, D11, D15** | Keep as-is | \<53% missing; clear inverse monotonic fraud pattern (low value = higher fraud); strong discriminative signal despite overlap in raw distributions |
| **D2** | High priority | Strongest inverse pattern — ratio D1/D10 of 0.10; likely important feature in SHAP analysis |
| **All D variables** | Missing → -999 | LightGBM sentinel strategy — these variables have structured missingness, so -999 is meaningful |
| **Inverse pattern** | Document | Values near 0 are associated with fraud; this is a strong behavioral signal worth monitoring in model |

### 5.7 Correlation Analysis

In this section, we examine linear relationships between numeric
variables to identify:

- Redundancy or multicollinearity
- Structural patterns and feature clusters
- Variables that may be useful (or harmful) in modeling

These insights can inform `feature selection`, `regularization`, and
`dimensionality reduction` strategies such as `PCA` or `Lasso`.

#### 5.7.1 Selection of Numerical Features

We restrict the analysis to numeric variables with fewer than 20%
missing values to ensure robustness of correlations.

```{r correlation-numeric-selection}
# Identify numeric columns with low NA ratio
# Exclude TransactionID explicitly
numeric_vars <- train |>
  select(where(is.numeric)) |>
  select(-TransactionID) |>
  summarise_all(~ mean(is.na(.))) |>
  pivot_longer(everything(), names_to = "Variable", values_to = "NA_Percent") |>
  filter(NA_Percent < 0.2) |>
  pull(Variable)

# Filter and clean
train_num <- train |>
  select(all_of(numeric_vars)) |>
  drop_na()
```

#### 5.7.2 Computing Correlation Matrix

We compute the **Pearson correlation matrix** for the cleaned numeric
dataset.

```{r correlation-matrix}
# Compute Pearson correlation matrix
cor_matrix <- cor(train_num, use = "complete.obs", method = "pearson")

# Quick check
dim(cor_matrix)
head(cor_matrix[, 1:5])
```

The full correlation matrix has 177 variables and is illegible as a
heatmap. Instead, we focus on two targeted views: correlations with the
target and multicollinearity among top predictors.

#### 5.7.3 Top Correlations with `isFraud`

We extract the variables most linearly associated with the target.

```{r correlation_heatmap2}
# Ordering by absolute correlation with isFraud
cor_target <- cor_matrix[, "isFraud"]
cor_target_sorted <- sort(abs(cor_target), decreasing = TRUE)

# View top correlations
head(cor_target_sorted, 20)
```

**Observation**:

- No feature has a strong linear correlation with isFraud (\> 0.15).
- The most correlated variables include V283, V281, V292, V291, V315,
  among others.
- This suggests that fraud detection may require nonlinear patterns or
  interactions.

#### 5.7.4 Correlation Among Top Predictors

We isolate and visualize the top 15 variables most correlated with the
target.

```{r correlation_heatmap2-subset}
# Obtain the top 15 variables with highest absolute correlation with isFraud (except isFraud itself)
top_vars <- names(cor_target_sorted)[2:16]

# Subset the correlation matrix for these variables
cor_subset <- cor_matrix[top_vars, top_vars]

# Heatmap
ggcorrplot(cor_subset,
  hc.order = TRUE, type = "lower", lab = TRUE,
  colors = c("#3B8BD4", "white", "#D85A30"),
  title = "Correlation Matrix of Top 15 Variables Correlated with isFraud",
  ggtheme = theme_minimal()
)
```

**Observation**:

- Variables such as `V282`–`V283`, `V313`–`V315`, and `V291`–`V292` show
  strong mutual correlations, indicating:
  - Possible feature redundancy
  - Opportunities for aggregation or dimensionality reduction

#### 5.7.5 Detecting Multicollinearity (Corr \> 0.90)

To identify highly redundant variables, we extract all variable pairs
with correlation above 0.90 (excluding self-correlation).

```{r correlation_heatmap2-subset2}
high_corr_pairs <- which(abs(cor_matrix) > 0.9 & abs(cor_matrix) < 1, arr.ind = TRUE)

# Remove symmetric duplicates
high_corr_df <- as.data.frame(high_corr_pairs)
high_corr_df <- high_corr_df[high_corr_df$row < high_corr_df$col, ]

# Add variable names and correlation values
high_corr_df <- high_corr_df |>
  mutate(
    var1 = rownames(cor_matrix)[row],
    var2 = colnames(cor_matrix)[col],
    corr = cor_matrix[cbind(row, col)]
  ) |>
  arrange(desc(abs(corr)))

head(high_corr_df, 10)
```

### Correlation Analysis Observations

- The top variables correlated with isFraud (e.g., `V283`, `V281`,
  `V292`, `V315`, `V289`) exhibit relatively low Pearson correlation
  coefficients (mostly under 0.15). This suggests that no single
  variable is linearly predictive of fraud on its own, reaffirming the
  complex and subtle nature of fraud detection.
- A subset of features (e.g., `V69`, `V90`, `V291`) show some moderate
  correlation, but these are likely to be more effective when used
  together or transformed (e.g., via non-linear models or interactions).
- Several pairs of variables exhibit very high mutual correlation (above
  0.99), such as:
  - `C7` ↔ `C12`
  - `C10` ↔ `C8`
  - `C1` ↔ `C11`
  - `V101` ↔ `V293`
  - `V95` ↔ `V101`, `V279`

These indicate redundancy or multicollinearity, which could impact model
stability and interpretability—particularly for linear models.

- `TransactionID` was excluded from the analysis, as it is an identifier
  and does not convey predictive information about the target.

#### 5.7.6 Pipeline Implications — Correlation

| Finding | Decision | Rationale |
|----|----|----|
| No linear feature has | None. Non linear relation detected. | r \< 0.15 with isFraud |
| V-clusters (V282–V283, V291–V292, V313–V315) show | Dimensional reduction | r \> 0.95 |
| C7 ↔ C12 and C8 ↔ C10 have | Dimensional reduction | r \> 0.99 |
| Correlation matrix computed with `complete.obs` | — | 1,478 rows used (0.25% of data) |

### 5.8 Feature Importance

Feature importance analysis is deferred to the Python modeling pipeline,
where it is computed using:

- **LightGBM built-in importance** (split and gain) — available directly
  from the trained model
- **SHAP values** — for per-instance interpretability and global feature
  ranking

The full dataset (with engineered features, sentinel imputation, and
categorical encoding) will be used for importance analysis, providing
more reliable rankings than an EDA-level Random Forest run on
complete-case data. See `configs/train.yaml` for SHAP configuration and
`src/models/train.py` for the implementation.

### 5.9 Cross-feature exploratory analysis

To better understand how fraud is distributed across categories, we
compute the fraud rate for selected categorical variables. This
highlights which categories are more likely to be associated with
fraudulent behavior.

```{r fraud-rate-function}
# Helper function to plot fraud rates for a categorical variable
plot_fraud_rate <- function(data, var) {
  data |>
    group_by(.data[[var]]) |>
    summarise(
      Count = n(),
      FraudRate = mean(isFraud, na.rm = TRUE)
    ) |>
    filter(Count > 200) |> # Optional: filter rare categories
    ggplot(aes(x = reorder(.data[[var]], -FraudRate), y = FraudRate)) +
    geom_col(fill = "#D85A30") +
    labs(
      title = paste("Fraud Rate by", var),
      x = var,
      y = "Fraud Rate"
    ) +
    theme_minimal()
}
```

#### 5.9.1 Fraud Rate by `ProductCD`

```{r fraud-rate-productcd}
# Plot for ProductCD
plot_fraud_rate(train, "ProductCD")
```

**Observations**:

- Category `C` has the highest fraud rate (nearly 12%), while `W` has
  the lowest.
- This suggests that transactions involving product type `C` should be
  treated with increased scrutiny during model training or rule design.

#### 5.9.2 Fraud Rate by `card4`

```{r fraud-rate-card4}
# Plot for card4
plot_fraud_rate(train, "card4")
```

**Observations**:

- `Discover` cards exhibit the highest fraud rate (around 8%), compared
  to others like `Visa` and `Mastercard.`
- This pattern may reflect underlying behavioral or systemic differences
  in transactions processed by these networks.

#### 5.9.3 Fraud Rate by `card6`

```{r fraud-rate-card6}
# Plot for card6
plot_fraud_rate(train, "card6")
```

**Observations**:

- `Credit` cards show a significantly higher fraud rate than debit
  cards.
- This might relate to the differing risk profiles or spending habits
  associated with each type.

#### 5.9.4 Summary

This cross-feature fraud rate analysis reinforces that some categorical
variables are informative for fraud prediction. Product type
(`ProductCD`), card provider (`card4`), and card type (`card6`) show
meaningful variation in fraud rates across categories. These features
may be especially valuable for downstream modeling, and further
interaction effects could be explored.

### 5.10 Dimensionality Reduction with PCA

Principal Component Analysis (PCA) is a technique used to reduce the
dimensionality of high-dimensional data by transforming correlated
variables into a smaller number of uncorrelated variables called
principal components. Although PCA is not ideal for interpretability, it
is useful for visualizing potential separation between fraudulent and
non-fraudulent transactions.

#### 5.10.1 Data Preparation for PCA

We use the cleaned numerical dataset from the correlation analysis
(`train_num`) and apply **z-score standardization** (mean = 0, std. dev.
= 1) to each variable before computing the principal components.

```{r pca-preparation}
# Ensure all numeric values are scaled
numeric_scaled <- train_num |>
  select(-isFraud) |>
  mutate_all(~ scale(.) |> as.vector())

# Attach target back
pca_data <- cbind(numeric_scaled, isFraud = train_num$isFraud)
```

**Note**:

- Out of 177 numeric features available (excluding `isFraud`), 176 were
  used in PCA.
- The variable `V305` was excluded because it had zero variance (i.e.,
  constant value across samples). Variables with zero variance do not
  contribute to PCA and must be removed.

#### 5.10.2 PCA Execution

To make PCA computationally tractable, we randomly sample 10,000
observations from the dataset and remove variables with zero variance.

```{r pca-execution}
# Step 1: sample 10,000 rows (with isFraud)
pca_sample <- train_num |>
  sample_n(10000)

# Step 2: drop isFraud and remove constant columns (zero variance)
scaled_sample <- pca_sample |>
  select(-isFraud) |>
  select(where(~ var(.x, na.rm = TRUE) > 0)) |>
  mutate_all(~ scale(.) |> as.vector())

# Step 3: apply PCA
pca_result <- prcomp(scaled_sample, center = TRUE, scale. = TRUE)

# Step 4: create plot data
pca_df <- as.data.frame(pca_result$x[, 1:2])
pca_df$isFraud <- factor(pca_sample$isFraud)

# Step 5: plot
ggplot(pca_df, aes(x = PC1, y = PC2, color = isFraud)) +
  geom_point(alpha = 0.5, size = 1) +
  labs(
    title = "PCA (Sample of 10K)",
    x = "PC1", y = "PC2",
    color = "Is Fraud"
  ) +
  theme_minimal()
```

Interpretation:

- Each point represents a transaction, projected into a 2D space formed
  by the two directions of greatest variance in the dataset: PC1 and
  PC2.
- Transactions are not clearly separated by fraud status, confirming
  that fraud patterns are subtle and not captured by linear combinations
  alone.
- Some outliers lie far along the principal axes, suggesting
  transactions that deviate significantly from typical behavior. These
  may be interesting candidates for anomaly detection.
- Most fraudulent transactions are embedded within the large cloud of
  normal transactions, again reinforcing the challenge of detecting
  fraud via simple linear boundaries.

#### 5.10.3 Variance Explained by Principal Components

We now assess how much variance each principal component captures.

```{r pca-variance-summary}
# Show variance explained by each component
summary(pca_result)
```

You can also visualize it more clarly:

```{r pca-variance-plot}
# Variance proportion
pca_var <- summary(pca_result)$importance[2, 1:10] # Proportion of variance

# Barplot
barplot(pca_var,
  names.arg = paste0("PC", 1:10),
  main = "Variance Explained by Top 10 Principal Components",
  xlab = "Principal Component",
  ylab = "Proportion of Variance",
  col = "#D85A30"
)
```

**Interpretation**:

- PC1 captures about 13–14% of the total variance, followed by PC2 with
  around 11%.
- The first 2 components together explain approximately 25% of the total
  variability.
- Additional components (PC3 to PC10) each add between 3% and 6%,
  showing that the variance is moderately distributed across many
  components.
- This implies that the data’s structure is complex and not dominated by
  a single axis of variation.

**Conclusion**:

PCA confirms that:

- Fraudulent transactions do not linearly separate from non-fraudulent
  ones in the first two components.
- However, PCA remains valuable for:
  - Anomaly detection and visual exploration,
  - Dimensionality reduction before clustering or linear modeling,
  - Reducing noise and multicollinearity in downstream tasks.

If used for modeling, a larger number of principal components should be
retained to preserve enough information. Future steps could include
training models on the top k components or integrating PCA into a
pipeline with clustering techniques.

### 5.11 Missing Data Analysis (NA Patterns)

While we’ve previously reviewed missing value percentages across
variables, visualizing their distribution across observations can reveal
structured patterns—such as whether certain blocks of features tend to
be missing together, or whether missingness is concentrated in specific
transaction types.

To do this, we generate a missing data heatmap of the top 50 variables
with the highest proportion of missing values. Due to computational
limits, we sample 5,000 rows from the full dataset.

```{r na-heatmap-setup}
# Step 1: select top 50 variables with highest NA percentage
na_top_vars <- train |>
  summarise(across(everything(), ~ mean(is.na(.)))) |>
  pivot_longer(everything(), names_to = "variable", values_to = "na_rate") |>
  arrange(desc(na_rate)) |>
  slice(1:50) |>
  pull(variable)
```

Now we have selected the top 50 variables with the highest missingness.
Next, we will visualize the missing data patterns.

```{r na-heatmap-plot}
# step 2: Sample for plotting
train_na_subset <- train |>
  select(all_of(na_top_vars)) |>
  slice_sample(n = 5000)

# step3: Plot heatmap with rotated labels
vis_miss(train_na_subset, sort_miss = TRUE) +
  labs(title = "Missing Data Heatmap - Top 50 Variables (Sample of 5,000 Rows)") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, size = 7),
    plot.title = element_text(size = 14, face = "bold")
  )
```

**Interpretation**:

- The majority of the top 50 variables exhibit high missingness, often
  above 85%, with some exceeding 95%.
- There is clear horizontal banding, which suggests that missingness is
  not completely random—it may be associated with particular transaction
  types or sources.
- Blocks of variables (e.g., many id\_ or V features) appear to be
  missing together, possibly reflecting system-level logging gaps or
  different data collection pipelines.

**Conclusion**:

This visual supports previous numeric summaries — missingness is
structural, driven by different data collection pipelines (transaction
vs identity tables).

#### 5.11.1 Pipeline Implications — Missing Data

| Finding | Decision | Rationale |
|----|----|----|
| High-missingness variables (\>80%) form distinct blocks | Drop D6–D9, D12–D14 per Section 5.6 | No structural explanation; non-missing values lack discriminative signal |
| id\_ variables nearly 100% missing without identity merge | Create `has_identity` binary flag | Captures structural missingness as a feature — 0.25x odds ratio per literature |
| V features with moderate missingness (\<20%) | Keep with sentinel imputation | 154 V features are core to model performance; -999 sentinel is standard for LightGBM |
| Missingness follows row-level patterns | No explicit missingness indicators per feature | Structural patterns are captured by `has_identity`; per-feature indicators add noise |

## 6. Summary and Next Steps

This EDA analyzed all available features in the IEEE-CIS Fraud Detection
dataset. The analysis focused on extracting actionable decisions for the
feature engineering pipeline rather than exhaustive description.

**Key Decisions:**

| Feature Group | Decision | Rationale |
|----|----|----|
| **TransactionAmt** | Keep, log-transform | Bimodal fraud distribution; extreme outliers carry fraud signal |
| **TransactionDT** | Keep, extract hour/DoW/Weekend | Temporal fraud patterns (early morning, weekday) |
| **ProductCD** | Keep, categorical | Product C: 12% fraud rate vs 3.5% baseline |
| **card1, card2, card3, card5** | Keep, frequency encoding | High cardinality; card aggregates capture velocity |
| **addr1, addr2** | Keep, frequency encoding | Geographic signal; addr1 is in top features |
| **M1–M9** | Keep, binary (T→1, F→0, NaN→-1) | Match flags — binary signal for identity verification |
| **P/R_emaildomain** | Keep, frequency encoding | Rare emails correlate with fraud |
| **C3** | **Drop** | 99.59% zeros; non-zero values have 0.21% fraud rate (vs 3.5% baseline) — inverse signal, no predictive value |
| **C4, C7, C8, C10, C12** | Keep (near-binary) | Value = 2 shows 10–14% fraud rate — clear discriminative signal |
| **C1, C2, C5, C6, C9, C11, C13, C14** | Keep (continuous) | Right-skewed; decile D10 shows elevated fraud; C13/C14 inverse pattern to monitor |
| **D1–D5, D10, D11, D15** | Keep | Clear inverse monotonic pattern: lower values = higher fraud (D2 ratio D1/D10 = 0.10) |
| **D6–D9** | **Drop** | 87–94% missing; no structural explanation |
| **D12–D14** | **Drop** | 89% missing; flat or non-monotonic patterns |
| **V features (154)** | Keep, apply correlation filter | Most correlated with target (V283, V281, V292); apply |
| **id features (38)** | Conditional on identity merge | Requires `has_identity` flag; sentinel imputation for missing |
| **has_identity** | **Engineered feature** | Binary flag capturing structural missingness from identity table |

**Architectural decisions for the Python pipeline:**

- **Missing value strategy**: LightGBM sentinel `-999` for numerical /
  `"missing"` for categorical — preserves structural missingness signal
  without explicit imputation
- **Encoding**: Label encoding for low-cardinality categoricals,
  frequency encoding for high-cardinality — avoids high-dimensional
  one-hot encoding
- **Feature importance**: Computed via LightGBM (split + gain) and SHAP
  in the Python training pipeline, not in EDA — ensures importance
  reflects engineered features
- **Model**: LightGBM baseline with Optuna HPO, evaluated on ROC-AUC and
  average precision (accuracy is useless at 3.5% fraud rate)

See `configs/features.yaml` for the full feature configuration and
`configs/train.yaml` for model and evaluation setup.
